---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
df = read.csv("activity.csv", header=TRUE)
```


## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day:

```{r}
dayTotal = aggregate(df$steps, list(df$date), sum, na.rm=TRUE)
head(dayTotal)
```

Make a histogram of the total number of steps taken each day.
Here, I first calculate the number of bars needed to get a good representation of the data:
```{r}
width = round(sqrt(length(dayTotal$x)))
bars = seq(min(dayTotal$x), max(dayTotal$x), by=( (max(dayTotal$x) - min(dayTotal$x)) / width ))
hist(dayTotal$x, breaks = bars, main = "Histogram of Total Number of Steps Taken per Day",
     xlab = "Total number of steps per day", col = "lightblue")
```

Calculate and report the mean and median of the total number of steps taken per day
(missing values are ignored):

```{r}
dayMean = aggregate(df$steps, list(df$date), mean, na.rm=TRUE)
dayMedian = aggregate(df$steps, list(df$date), median, na.rm=TRUE)
```

Below is the output for the mean number of steps taken per day.

```{r}
data.frame(Date = dayMean$Group.1, Mean = dayMean$x)
```

And here is the output for the median number of steps:
```{r}
data.frame(Date = dayMedian$Group.1, Median = dayMedian$x)
```


## What is the average daily activity pattern?

First, I calculate the mean number of steps for each day interval:
```{r}
intervals = as.factor(df$interval)
intervalMean = aggregate(df$steps, list(intervals), mean, na.rm=TRUE)
colnames(intervalMean) = c("Interval", "Mean")
```

Then, I create the time series plot of the 5-minute interval (x-axis) and 
the average number of steps taken, averaged across all days (y-axis):
```{r}
plot(intervalMean$Mean ~ intervalMean$Interval,
     main = paste("Time Series Plot of Average Number of Steps Taken\n",
                  "At Each 5-Minute Daily Interval"),
     xlab = "Daily Interval (in 5-minute increments)",
     ylab = "Mean Number of Steps Taken")
lines(intervalMean$Mean)
```

#### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
intervalMean[which(intervalMean$Mean == max(intervalMean$Mean)),]
```
In other words, interval 835 with mean 206.1698 steps.

## Inputing missing values
Calculate and report the total number of missing values in the dataset:
```{r}
length(which(is.na(df$steps) == TRUE))
```

Devise a strategy for filling in all of the missing values in the dataset.
Here, I replace NAs with the mean value for the specific time interval:
```{r}
newSteps = df$steps
for (i in which(is.na(newSteps) == TRUE)) {
  newSteps[i] = intervalMean[which(intervalMean$Interval == df$interval[i]),2]
}
```

Create a new dataset that is equal to the original dataset but with the missing data filled in:
```{r}
newDf = data.frame(newSteps, df$date, df$interval)
colnames(newDf) = c("steps", "date", "interval")
```

Make a histogram of the total number of steps taken each day:
```{r}
newDayTotal = aggregate(newDf$steps, list(newDf$date), sum, na.rm=TRUE)
hist(newDayTotal$x, breaks = bars, main = "Histogram of Total Number of Steps Taken per Day",
     xlab = "Total number of steps per day", col = "lightpink")
```

Calculate and report the mean and median total number of steps taken per day:
```{r}
newDayMean = aggregate(newDf$steps, list(newDf$date), mean, na.rm=TRUE)
newDayMedian = aggregate(newDf$steps, list(newDf$date), median, na.rm=TRUE)
```
Below is the output for the means:
```{r}
data.frame(Date = newDayMean$Group.1, Mean = newDayMean$x)
```
And this is the output for the medians: 
```{r}
data.frame(Date = newDayMedian$Group.1, Median = newDayMedian$x)
```

#### Do these values differ from the estimates from the first part of the assignment? 
Both the mean and the median differ from the estimates where there were previously NA values. 

#### What is the impact of imputing missing data on the estimates of the total daily number of steps?
Since the NAs have been filled with the mean values, there is a greater frequency of totals at
the center, resulting in a more bell-shaped distribution for the total step count. 

## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable in the dataset with two levels - "weekday" and "weekend":
```{r}
weekDay = weekdays(as.POSIXct(df$date))
newDf$dayType = as.factor(ifelse( (weekDay == "Saturday" | weekDay == "Sunday"),
                                 "weekend", "weekday"))
```

Calculating the interval means for weekdays and weekends:
```{r}
newIntervalMean = aggregate(newDf$steps, list(intervals,
                                              newDf$dayType == "weekend"), mean, na.rm=TRUE)
colnames(newIntervalMean) = c("Interval", "isWeekend", "Mean")
intMeanDay = newIntervalMean[newIntervalMean$isWeekend == FALSE, c(1,3)]
intMeanWeek = newIntervalMean[newIntervalMean$isWeekend == TRUE, c(1,3)]
```

Finally, here is the panel plot containing a time series plot of the 5-minute interval (x-axis) 
and the average number of steps taken, averaged across all weekday days or weekend days (y-axis):
```{r}
par(mfrow=c(2,1))
plot(intMeanDay$Mean ~ intMeanDay$Interval,
     xlab = "Daily Interval (in 5-minute increments)",
     ylab = "Mean Number of Steps Taken")
lines(intMeanDay$Mean)
title(paste("Average Number of Steps Taken\n",
            "At Each 5-Minute Daily Interval (Weekdays)"))
plot(intMeanWeek$Mean ~ intMeanWeek$Interval,
     xlab = "Daily Interval (in 5-minute increments)",
     ylab = "Mean Number of Steps Taken",
     ylim = c(0,200))
lines(intMeanWeek$Mean)
title(paste("Average Number of Steps Taken\n",
            "At Each 5-Minute Daily Interval (Weekends)"))

```


